---
title: "USA Unhealthy Weather Events (1950-2011)"
author: "David Dessert"
date: "Tuesday, February 17, 2015"
output: html_document
keep_md: true
---

## Synopsis
This report identifies weather severe events harmful to people and/or property. The analysis uses historical data from the [NOAA Storm Database](http://www.ncdc.noaa.gov/stormevents/) between 1950 and 2011.

## Load necessary packages to execute
```{r}
library(ggplot2)
library(plyr)
```

## R Software environment
```{r}
sessionInfo()
```

## Download the source file
```{r}
inFile <- "./data/repdata_data_StormData.csv.bz2"
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              inFile)
```

## Load and preprocessing the data
The input data file is stored in the ./data subdirectory and
is in a comma-separated variable format (.csv extension) within 
a zipped file (.bz2 extension). Extracting and loading takes a while.
```{r}
data <- read.csv(bzfile(inFile))
df <- data[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP")]
summary(df)
```
Extract the year of occurrence
```{r}
df$YEAR <- as.numeric(format(strptime(data$BGN_DATE, format="%m/%d/%Y %H:%M:%S"), "%Y"))
```
Combine EVTYPE's that are identical (ignore case difference)
```{r}
levels(df$EVTYPE) <- toupper(levels(df$EVTYPE))
```

Merge some EVTYPE's into the event types enumerated by the NWS.  
The people who entered the data into the database were not very consistent.
Here I coerce the various EVTYPE's used into one of the 48 EVTYPE's
spelled out specifically in the [Storm Event documentation](http://www.nws.noaa.gov/directives/).  

Event Type(s)     | Dates Available
----------------- | --------------------
Tornado           | Jan 1950 - Present
Thunderstorm Wind | Jan 1955 - Present
Hail              | Jan 1955 - Present
All others        | Jan 1996 - Present

```{r EVTYPE_cleanup}
# For some reason, DEBRIS FLOW is not an entry. Add it to our levels.
levels(df$EVTYPE) <- c(levels(df$EVTYPE), "DEBRIS FLOW")

df$EVTYPE[grep("AVAL", df$EVTYPE)] <- "AVALANCHE"
df$EVTYPE[grep("BLIZ", df$EVTYPE)] <- "BLIZZARD"
df$EVTYPE[grep("COAST.*FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("CSTL FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("BEACH FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("TIDAL FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COASTALSTORM", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COASTAL STORM", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("UNSEASONABLY COOL", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("HYPERTHERMIA", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("WIND CH", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("ICE.*FOG", df$EVTYPE)] <- "FREEZING FOG"
df$EVTYPE[grep("FRE.*FOG", df$EVTYPE)] <- "FREEZING FOG"
df$EVTYPE[grep("MUD.*SL", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("ROCK SLIDE", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("LANDSL", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("COASTAL EROSION", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("DAM BREAK", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("DENS.*FOG", df$EVTYPE)] <- "DENSE FOG"
df$EVTYPE[grep("^FOG$", df$EVTYPE)] <- "DENSE FOG"
df$EVTYPE[grep("SMOKE", df$EVTYPE)] <- "DENSE SMOKE"
df$EVTYPE[grep("DROU", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("EXCESSIVELY DRY", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("UNSEASONABLY DRY", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("HOT/DRY PATTERN", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("WARM DRY CONDITIONS", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("RECORD DRYNESS", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("DUST.*DEV", df$EVTYPE)] <- "DUST DEVIL"
df$EVTYPE[grep("DUST.*ST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("BLOWING DUST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("SAHARAN DUST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("HEAT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("UNSEASONABLY WARM & WET", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("UNSEASONABLY HOT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("CHILL", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("HYPOTHERMIA", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("LOW TEMP", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("FLASH", df$EVTYPE)] <- "FLASH FLOOD"
df$EVTYPE[grep("^FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("FROST", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("FREEZE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("BLACK ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("PATCHY ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("GLAZE ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("FUNNEL", df$EVTYPE)] <- "FUNNEL CLOUD"
df$EVTYPE[grep("ROTATING WALL CLOUD", df$EVTYPE)] <- "FUNNEL CLOUD"
df$EVTYPE[grep("HAIL", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("ICE PELLETS", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("HEAT", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("^UNSEASONABLY WARM$", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("^UNSEASONABLY WARM YEAR$", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("RAIN DAMAGE", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXTREMELY WET", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAV.* SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("RECORD.* SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOWSTORM", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOW.* RECORD", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("FIRST SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("EXCESSIVE SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("EARLY SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("^RAIN/SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SURF", df$EVTYPE)] <- "HIGH SURF"
df$EVTYPE[grep("^HIGH SEAS", df$EVTYPE)] <- "HIGH SURF"
df$EVTYPE[grep("HURRICANE", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("REMNANTS OF FLOYD", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("TYPHOON", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("ICE ST", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("ICE FLOES", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("ICE JAM", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("IC[EY].* ROAD", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("LAKE FL", df$EVTYPE)] <- "LAKESHORE FLOOD"
df$EVTYPE[grep("LAKE SN", df$EVTYPE)] <- "LAKE-EFFECT SNOW"
df$EVTYPE[grep("LIGHTN", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("LIGNTNING", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("LIGHTING", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("MARINE T.* WIND", df$EVTYPE)] <- "MARINE THUNDERSTORM WIND"
df$EVTYPE[grep("ROUGH SEAS", df$EVTYPE)] <- "MARINE HIGH WIND"
df$EVTYPE[grep("MARINE MISHAP", df$EVTYPE)] <- "MARINE HIGH WIND"
df$EVTYPE[grep("RIP .*CURR", df$EVTYPE)] <- "RIP CURRENT"
df$EVTYPE[grep("SLEET", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("FREEZING RAIN", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("FREEZING DRIZZLE", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("^HIGH WAVES", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("SURGE", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("HIGH TIDES", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("ASTRONOMICAL HIGH TIDE", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("STR.* WIND", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND GUST", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND STORM", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND ADVIS", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("GUST.*WIND", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WIND$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WINDS$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WIND DAMAGE$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("THU.*WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTM WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("DRY MICROBURST WINDS", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTM WND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTMW", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORMW", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORM WIN", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORM W INDS", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TUNDERSTORM WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("^SEVERE THUNDERSTORM", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("GUSTNADO", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TORN", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("WHIRLWIND", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("TROP", df$EVTYPE)] <- "TROPICAL STORM"
df$EVTYPE[grep("VOLC", df$EVTYPE)] <- "VOLCANIC ASH"
df$EVTYPE[grep("SPOU", df$EVTYPE)] <- "WATERSPOUT"
df$EVTYPE[grep("FIRE", df$EVTYPE)] <- "WILDFIRE"
df$EVTYPE[grep("WINTER S", df$EVTYPE)] <- "WINTER STORM"
df$EVTYPE[grep("WINTER W", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("SNOW SQUALLS", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("WINT.* MIX", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^ICE$", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^ICE/SNOW$", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW AND ICE", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW/BLOWING SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^LIGHT SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW/FREEZING RAIN", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^DRIFTING SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW SQUALL", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^THUNDERSNOW SHOWER", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^MIXED PRECIP", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW FREEZING RAIN", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^LAKE[ -]EFFECT SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^MIXED PRECIPITATION", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("HEAVY RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXCESSIVE RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RAINSTORM", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RAIN \\(HEAVY\\)", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("TORR.* RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("PROLONGED RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HVY RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAVY PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAVY SHOWER", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXCESSIVE PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RECORD PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^THUNDERSTORM$", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^WET YEAR", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HIGH +WIND", df$EVTYPE)] <- "HIGH WIND"         # only wind, no storms
df$EVTYPE[grep("^STORM FORCE WINDS", df$EVTYPE)] <- "HIGH WIND"         # only wind, no storms
df$EVTYPE[grep("RIVER FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("STREAM FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("URBAN.* FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("RURAL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("STREET FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HIGHWAY FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SNOW.* FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("MINOR FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("LOCAL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SMALL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("MAJOR FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("BREAKUP FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("ICE JAM FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SML STREAM FLD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("DAM FAILURE", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HIGH WATER", df$EVTYPE)] <- "FLOOD"
```

## Isolate important EVTYPE's by non-zero FATALITIES, INJURIES, or PROPDMG
```{r}
x <- ddply(df, .(EVTYPE), summarize, imp_events = (FATALITIES>0) || (INJURIES>0) || (PROPDMG>0))
x$EVTYPE[x$imp_events]
```

## Integrate the units for Property Damage
The suffixes used in the database are varied and I'm unsure what the
people who keyed them in really meant, but here's what I'm assuming:  
Table: Property Damage Multipliers
PROPDMGEXP | Multiplier
---------- | ------------------
b, B       | 10^9^ (Billions)
m, M       | 10^6^ (Millions)
k, K       | 10^3^ (Thousands)
h, H       | 10^2^ (Hundreds)
0..8       | 10^1^ (Tens)
+          | 10^0^ (Ones)
-          | 10^0^ (Ones)
?          | 10^0^ (Ones)
" "        | 10^0^ (Ones)

```{r}
m <- rep(1, length = nrow(df))
m[toupper(df$PROPDMGEXP)=="B"] <- 1E9
m[toupper(df$PROPDMGEXP)=="M"] <- 1E6
m[toupper(df$PROPDMGEXP)=="K"] <- 1E3
m[toupper(df$PROPDMGEXP)=="H"] <- 1E2
df$PROPDMG <- df$PROPDMG * m
```

## Find the Top-10 in each category:
Fatalities
Injuries
Property Damage
Count of events
```{r}
x <- ddply(df, .(EVTYPE, YEAR), summarize, 
           count=length(EVTYPE), 
           fatal.sum=sum(FATALITIES), 
           injury.sum=sum(INJURIES), 
           prop_dmg.sum=sum(PROPDMG)
           )
```
Top-N Ordered by property damage per event
```{r}
N <- 4      # Number of EVTYPE's to plot
# Note that property damage was not reported until 1993, so the
# maximum number of years is limited to 19 (1993 - 2011)
y <- ddply(x, .(EVTYPE), summarize, 
           count=min(length(YEAR), 19),
           damage=sum(prop_dmg.sum), 
           damage_per_year=sum(prop_dmg.sum)/min(length(YEAR), 19))
top_damage <- head(y[order(y$damage_per_year, decreasing=TRUE),], 6)
p <- ggplot(x[x$EVTYPE %in% top_damage$EVTYPE & (x$YEAR >= 1993),], aes(x=EVTYPE, y=prop_dmg.sum)) 
p <- p + geom_violin(alpha=1/4, aes(fill=EVTYPE)) 
p <- p + geom_jitter(aes(colour=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"))
p <- p + xlab('Storm Event')                        # x-axis label
p <- p + ylab('Yearly Property Damage ($)')             # y-axis label
p <- p + ggtitle("Top Storm Property Hazards")      # graph title
print(p)
```

Top-N Ordered by fatalities & injuries per year
```{r}
y <- ddply(x, .(EVTYPE), summarize, 
           count=length(YEAR), 
           fatal=sum(fatal.sum), 
           fatal_per_year=sum(fatal.sum)/length(YEAR),
           injury=sum(injury.sum),
           injury_per_year=sum(injury.sum)/length(YEAR))
top_health <- unique(rbind(head(y[order(y$fatal_per_year, decreasing=TRUE),], N), head(y[order(y$injury_per_year, decreasing=TRUE),], N)))
# Plot with clustering
p <- ggplot(x[x$EVTYPE %in% top_health$EVTYPE,], aes(x=fatal.sum, y=injury.sum, colour=EVTYPE)) + geom_point() + stat_ellipse(geom='polygon', alpha=1/8, aes(fill=EVTYPE))
# plot the average for all years
# p <- p + geom_point(data=top_health, aes(x=fatal_per_year, y=injury_per_year)) + geom_point(aes(colour=EVTYPE))
p <- p + xlab('Fatalities / Year')              # x-axis label
p <- p + ylab('Injuries / Year')                # y-axis label
p <- p + ggtitle("Top Storm Health Hazards")        # graph title
print(p)
```

