---
title: "USA Hazardous Weather Events (1950-2011)"
author: "David Dessert"
date: "Tuesday, February 17, 2015"
output:
  html_document:
    fig_width: 9
keep_md: yes
---

## Synopsis
This report identifies weather severe events most harmful to people and property. Using historical data from the [NOAA Storm Database](http://www.ncdc.noaa.gov/stormevents/) between 1950 and 2011, annualized fatalities and injuries counts of weather events were analyzed and the worst weather events identified. By our analysis, tornadoes and heat events caused the most annual fatalities and injuries while droughts, floods, and hurricanes caused the most crop and property damage.

## Data Processing
First, we load the necessary packages to execute
```{r}
library(ggplot2)
library(plyr)
```

This is the R Software environment we are using for this analysis.
```{r}
sessionInfo()
```

The input data file is downloaded from the Coursera web site and stored in the ./data subdirectory. 
The source file is in a comma-separated variable format (.csv extension) within 
a zipped file (.bz2 extension). Extracting and loading takes a while.
```{r}
inFile <- "./data/repdata_data_StormData.csv.bz2"
# download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
#               inFile)
data <- read.csv(bzfile(inFile))
df <- data[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
summary(df)
```

### Cleaning the Data
Not all weather events were available for every year of the database.
This table indicates the years when weather events were available.

| Event Type(s)     | Dates Available     |
| ----------------- | ------------------- |
| Tornado           | Jan 1950 - Present  |
| Thunderstorm Wind | Jan 1955 - Present  |
| Hail              | Jan 1955 - Present  |
| All others        | Jan 1996 - Present  |

Additionally, damage estimates were not recorded in the database prior to 1993.
Extract the year of occurrence from the BGN_DATE field. We use this field to
keep track of the year the event occurred.
```{r year_extract}
df$YEAR <- as.numeric(gsub("\\d+/\\d+/(\\d+).*", "\\1", data$BGN_DATE))
```

Combine weather event names (EVTYPE) that are identical except for upper/lower case differences.
```{r}
levels(df$EVTYPE) <- toupper(levels(df$EVTYPE))
```
Merge weather event names into the event types enumerated by the NWS.
The people who entered the data into the database were not very consistent.
Here I coerce the various weather events used into one of the 48 weather events
spelled out specifically in the [Storm Event documentation](http://www.nws.noaa.gov/directives/sym/pd01016005curr.pdf).
Not all weather events remaining will have been coerced into the correct types,
but all weather events large enough to affect the analysis are corrected.

```{r EVTYPE_cleanup}
# For some reason, DEBRIS FLOW is not an entry. Add it to our levels (at the end).
levels(df$EVTYPE) <- c(levels(df$EVTYPE), "DEBRIS FLOW")

# Start coercion of EVTYPE's into the 48 "approved" types
df$EVTYPE[grep("AVAL", df$EVTYPE)] <- "AVALANCHE"
df$EVTYPE[grep("BLIZ", df$EVTYPE)] <- "BLIZZARD"
df$EVTYPE[grep("COAST.*FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("CSTL FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("BEACH FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("TIDAL FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COASTALSTORM", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COASTAL STORM", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("UNSEASONABLY COOL", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("HYPERTHERMIA", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("WIND CH", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("ICE.*FOG", df$EVTYPE)] <- "FREEZING FOG"
df$EVTYPE[grep("FRE.*FOG", df$EVTYPE)] <- "FREEZING FOG"
df$EVTYPE[grep("MUD.*SL", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("ROCK SLIDE", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("LANDSL", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("COASTAL EROSION", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("DAM BREAK", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("DENS.*FOG", df$EVTYPE)] <- "DENSE FOG"
df$EVTYPE[grep("^FOG$", df$EVTYPE)] <- "DENSE FOG"
df$EVTYPE[grep("SMOKE", df$EVTYPE)] <- "DENSE SMOKE"
df$EVTYPE[grep("DROU", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("EXCESSIVELY DRY", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("UNSEASONABLY DRY", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("HOT/DRY PATTERN", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("WARM DRY CONDITIONS", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("RECORD DRYNESS", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("DUST.*DEV", df$EVTYPE)] <- "DUST DEVIL"
df$EVTYPE[grep("DUST.*ST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("BLOWING DUST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("SAHARAN DUST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("HEAT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("UNSEASONABLY WARM & WET", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("UNSEASONABLY HOT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("CHILL", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("HYPOTHERMIA", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("LOW TEMP", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("FLASH", df$EVTYPE)] <- "FLASH FLOOD"
df$EVTYPE[grep("^FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("FROST", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("FREEZE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("BLACK ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("PATCHY ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("GLAZE ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("FUNNEL", df$EVTYPE)] <- "FUNNEL CLOUD"
df$EVTYPE[grep("ROTATING WALL CLOUD", df$EVTYPE)] <- "FUNNEL CLOUD"
df$EVTYPE[grep("HAIL", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("ICE PELLETS", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("HEAT", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("^UNSEASONABLY WARM$", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("^UNSEASONABLY WARM YEAR$", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("RAIN DAMAGE", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXTREMELY WET", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAV.* SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("RECORD.* SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOWSTORM", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOW.* RECORD", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("FIRST SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("EXCESSIVE SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("EARLY SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("^RAIN/SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SURF", df$EVTYPE)] <- "HIGH SURF"
df$EVTYPE[grep("^HIGH SEAS", df$EVTYPE)] <- "HIGH SURF"
df$EVTYPE[grep("HURRICANE", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("REMNANTS OF FLOYD", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("TYPHOON", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("ICE ST", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("ICE FLOES", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("ICE JAM", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("IC[EY].* ROAD", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("LAKE FL", df$EVTYPE)] <- "LAKESHORE FLOOD"
df$EVTYPE[grep("LAKE SN", df$EVTYPE)] <- "LAKE-EFFECT SNOW"
df$EVTYPE[grep("LIGHTN", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("LIGNTNING", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("LIGHTING", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("MARINE T.* WIND", df$EVTYPE)] <- "MARINE THUNDERSTORM WIND"
df$EVTYPE[grep("ROUGH SEAS", df$EVTYPE)] <- "MARINE HIGH WIND"
df$EVTYPE[grep("MARINE MISHAP", df$EVTYPE)] <- "MARINE HIGH WIND"
df$EVTYPE[grep("RIP .*CURR", df$EVTYPE)] <- "RIP CURRENT"
df$EVTYPE[grep("SLEET", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("FREEZING RAIN", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("FREEZING DRIZZLE", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("^HIGH WAVES", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("SURGE", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("HIGH TIDES", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("ASTRONOMICAL HIGH TIDE", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("STR.* WIND", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND GUST", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND STORM", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND ADVIS", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("GUST.*WIND", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WIND$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WINDS$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WIND DAMAGE$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("THU.*WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTM WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("DRY MICROBURST WINDS", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTM WND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTMW", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORMW", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORM WIN", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORM W INDS", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TUNDERSTORM WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("^SEVERE THUNDERSTORM", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("GUSTNADO", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TORN", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("WHIRLWIND", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("TROP", df$EVTYPE)] <- "TROPICAL STORM"
df$EVTYPE[grep("VOLC", df$EVTYPE)] <- "VOLCANIC ASH"
df$EVTYPE[grep("SPOU", df$EVTYPE)] <- "WATERSPOUT"
df$EVTYPE[grep("FIRE", df$EVTYPE)] <- "WILDFIRE"
df$EVTYPE[grep("WINTER S", df$EVTYPE)] <- "WINTER STORM"
df$EVTYPE[grep("WINTER W", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("SNOW SQUALLS", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("WINT.* MIX", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^ICE$", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^ICE/SNOW$", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW AND ICE", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW/BLOWING SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^LIGHT SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW/FREEZING RAIN", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^DRIFTING SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW SQUALL", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^THUNDERSNOW SHOWER", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^MIXED PRECIP", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW FREEZING RAIN", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^LAKE[ -]EFFECT SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^MIXED PRECIPITATION", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("HEAVY RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXCESSIVE RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RAINSTORM", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RAIN \\(HEAVY\\)", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("TORR.* RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("PROLONGED RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HVY RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAVY PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAVY SHOWER", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXCESSIVE PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RECORD PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^THUNDERSTORM$", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^WET YEAR", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HIGH +WIND", df$EVTYPE)] <- "HIGH WIND"         # only wind, no storms
df$EVTYPE[grep("^STORM FORCE WINDS", df$EVTYPE)] <- "HIGH WIND"         # only wind, no storms
df$EVTYPE[grep("RIVER FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("STREAM FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("URBAN.* FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("RURAL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("STREET FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HIGHWAY FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SNOW.* FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("MINOR FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("LOCAL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SMALL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("MAJOR FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("BREAKUP FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("ICE JAM FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SML STREAM FLD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("DAM FAILURE", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HIGH WATER", df$EVTYPE)] <- "FLOOD"
```

### Incorporate the units into Crop & Property Damages
The suffixes used in the database are quite varied. Based on a review of the
entries, here is what I'm assuming each multiplier means:  

Table: Damage Multipliers

| DMGEXP     | Multiplier   | Interpretation           |
| ---------- | ------------:| ------------------------ |
| b, B       | 10^9^        | Billions                 |
| m, M       | 10^6^        | Millions                 |
| k, K       | 10^3^        | Thousands                |
| h, H       | 10^2^        | Hundreds                 |
| 0..8       | 10^1^        | extra significant digit  |
| +          | 10^0^        | >                        |
| -          | 10^0^        | <                        |
| ?          | 10^0^        | usually 0                |
| " "        | 10^0^        | usually 0                |

Fix up the property and crop damage estimates using the multipliers based on the information in the table above.
```{r cost_fixups, tidy=TRUE}
m <- rep(1, length = nrow(df))                              # default to 1
m[toupper(df$PROPDMGEXP)=="B"] <- 1E9                       # B, b
m[toupper(df$PROPDMGEXP)=="M"] <- 1E6                       # M, m
m[toupper(df$PROPDMGEXP)=="K"] <- 1E3                       # K, k
m[toupper(df$PROPDMGEXP)=="H"] <- 1E2                       # H, h
m[df$PROPDMGEXP %in% c('0','1','2','3','4','5','6','7','8','9')] <- 10                  # 0-9
df$PROPDMG <- df$PROPDMG * m

m <- rep(1, length = nrow(df))                              # default to 1
m[toupper(df$CROPDMGEXP)=="B"] <- 1E9                       # B, b
m[toupper(df$CROPDMGEXP)=="M"] <- 1E6                       # M, m
m[toupper(df$CROPDMGEXP)=="K"] <- 1E3                       # K, k
m[toupper(df$CROPDMGEXP)=="H"] <- 1E2                       # H, h
m[df$CROPDMGEXP %in% c('0','1','2','3','4','5','6','7','8','9')] <- 10                  # 0-9
df$CROPDMG <- df$CROPDMG * m
```
Tabulate the weather events by {EVTYPE, YEAR} into a data frame that
we'll use for the rest of the summary. In this report, we are annualizing
the weather events. Tornadoes have been reported since 1950 whereas most
weather events have ony been reported since 1993, so using a total count
for comparison would skew the results.
```{r}
N <- 4      # Number of top EVTYPE's to be plotted in each category
df_summary <- ddply(df, .(EVTYPE, YEAR), summarize, 
           count=length(EVTYPE), 
           fatal.sum=sum(FATALITIES), 
           injury.sum=sum(INJURIES), 
           prop_dmg.sum=sum(PROPDMG),
           crop_dmg.sum=sum(CROPDMG)
           )
```
## Results

For each category of fatalities, injuries, crop damage, and property damage, 
we'll be getting the top `r N` weather events for display. 
When displaying the fatalities/injuries figure, less than `r 2*N` weather events 
are displayed because of overlap in the fataility top `r N` weather events 
and injury top `r N` weather events. The same comment holds true for the
crop/property damage figure.  

### Property Damage

This figure shows summarizes the annual damage reports for the Top 6 weather events
that cause property damage. Each dot represents the damage caused in a particular
year and the shape's width corresponds to more years with the damage amount.
```{r}
# Note that property damage was not reported until 1993, so the
# maximum number of years is limited to 19 (1993 - 2011)
df_figdata <- ddply(df_summary, .(EVTYPE), summarize, 
           count=min(length(YEAR), 19),
           damage=sum(prop_dmg.sum), 
           damage_per_year=sum(prop_dmg.sum)/min(length(YEAR), 19))
top_damage <- head(df_figdata[order(df_figdata$damage_per_year, decreasing=TRUE),], 6)
```

```{r warning=FALSE, fig.width=9, fig.cap="Weather-Related Property Damages"}
p <- ggplot(df_summary[df_summary$EVTYPE %in% top_damage$EVTYPE & (df_summary$YEAR >= 1993),], aes(x=EVTYPE, y=prop_dmg.sum)) 
p <- p + geom_violin(alpha=1/4, aes(fill=EVTYPE)) 
p <- p + geom_jitter(aes(colour=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"))
p <- p + xlab('Storm Event')                        # x-axis label
p <- p + ylab('Annual Property Damage ($)')             # y-axis label
p <- p + ggtitle("Top Weather Property Hazards")      # graph title
print(p)
```

Table: Property Damage Summary

| Rank | Weather Event    | Total Property Damage | Annualized Property Damage |
| ---- | ---------- | ------------------:| ------------------------:|
|   1  | `r top_damage$EVTYPE[1]` | $ `r top_damage$damage[1]/1E9`B   | $ `r top_damage$damage_per_year[1]/1E9`B |
|   2  | `r top_damage$EVTYPE[2]` | $ `r top_damage$damage[2]/1E9`B   | $ `r top_damage$damage_per_year[2]/1E9`B |
|   3  | `r top_damage$EVTYPE[3]` | $ `r top_damage$damage[3]/1E9`B   | $ `r top_damage$damage_per_year[3]/1E9`B |
|   4  | `r top_damage$EVTYPE[4]` | $ `r top_damage$damage[4]/1E9`B   | $ `r top_damage$damage_per_year[4]/1E9`B |
|   5  | `r top_damage$EVTYPE[5]` | $ `r top_damage$damage[5]/1E9`B   | $ `r top_damage$damage_per_year[5]/1E9`B |
|   6  | `r top_damage$EVTYPE[6]` | $ `r top_damage$damage[6]/1E9`B   | $ `r top_damage$damage_per_year[6]/1E9`B |

### Fatality and Injury Summary

This figure shows summarizes past weather events recording the top `r N`
annualized fatality and top `r N` injury counts. Each dot represents the
recorded fatalities/injuries in a particular year. For each weather event,
an oval was drawn to encompass a 95% CI illustrating where future years' 
expected fatalities/injuries counts may lie.
```{r warning=FALSE, fig.width=9}
df_figdata <- ddply(df_summary, .(EVTYPE), summarize, 
           count=length(YEAR), 
           fatal=sum(fatal.sum), 
           fatal_per_year=sum(fatal.sum)/length(YEAR),
           injury=sum(injury.sum),
           injury_per_year=sum(injury.sum)/length(YEAR))
top_fatal <- df_figdata[order(df_figdata$fatal_per_year, decreasing=TRUE),]
top_injury <- df_figdata[order(df_figdata$injury_per_year, decreasing=TRUE),]
top_health <- unique(rbind(head(top_fatal, N), 
                           head(top_injury, N)))
# Plot with clustering
p <- ggplot(df_summary[df_summary$EVTYPE %in% top_health$EVTYPE,], aes(x=fatal.sum, y=injury.sum, colour=EVTYPE)) + geom_point() + stat_ellipse(geom='polygon', alpha=1/8, aes(fill=EVTYPE))
# For each EVTYPE, add a larger point to indicate the average for all years
# p <- p + geom_point(data=top_health, aes(x=fatal_per_year, y=injury_per_year)) + geom_point(aes(colour=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 10, 100, 1000, 10000), labels=c("1","10","10","1000","10K"), limits=c(0.9, 10000))
p <- p + scale_x_log10(breaks=c(1, 10, 100, 1000, 10000), labels=c("1","10","10","1000","10K"), limits=c(0.9, 10000))
p <- p + xlab('Annual Fatalities')                    # x-axis label
p <- p + ylab('Annual Injuries')                      # y-axis label
p <- p + ggtitle("Top Weather Health Hazards")        # graph title
print(p)
```

The results indicate that Heat and Tornadoes are the major weather-related causes of injuries and fatalities. Floods, Flash Floods, and Lightning are major causes of injuries, but not fatalities.

Table: Fatality / Injury Summary

| Rank | Fatal Weather Events | Total / Annual Fatalities | Injury Weather Events | Total / Annual Injuries 
| ---- | ---------- | ------------------:| ---------- | ------------------------:|
| 1 | `r top_fatal$EVTYPE[1]` | `r top_fatal$fatal[1]` / `r round(top_fatal$fatal_per_year[1])` | `r top_injury$EVTYPE[1]` | `r top_injury$injury[1]` / `r round(top_injury$injury_per_year[1])` |
| 2 | `r top_fatal$EVTYPE[2]` | `r top_fatal$fatal[2]` / `r round(top_fatal$fatal_per_year[2])` | `r top_injury$EVTYPE[2]` | `r top_injury$injury[2]` / `r round(top_injury$injury_per_year[2])` |
| 3 | `r top_fatal$EVTYPE[3]` | `r top_fatal$fatal[3]` / `r round(top_fatal$fatal_per_year[3])` | `r top_injury$EVTYPE[3]` | `r top_injury$injury[3]` / `r round(top_injury$injury_per_year[3])` |
| 4 | `r top_fatal$EVTYPE[4]` | `r top_fatal$fatal[4]` / `r round(top_fatal$fatal_per_year[4])` | `r top_injury$EVTYPE[4]` | `r top_injury$injury[4]` / `r round(top_injury$injury_per_year[4])` |

### Property and Crop Damage Summary

This figure shows summarizes past weather events recording the top `r N`
annualized property and top `r N` crop damage dollar amounts. Each dot represents the
recorded property/crop damage in a particular year. For each weather event,
an oval was drawn to encompass a 95% CI illustrating where future years' 
expected property/crop damages may lie.
```{r warning=FALSE, fig.width=9}
df_figdata <- ddply(df_summary, .(EVTYPE), summarize, 
           count=length(YEAR), 
           prop=sum(prop_dmg.sum), 
           prop_per_year=sum(prop_dmg.sum)/length(YEAR),
           crop=sum(crop_dmg.sum),
           crop_per_year=sum(crop_dmg.sum)/length(YEAR))
top_prop <- df_figdata[order(df_figdata$prop_per_year, decreasing=TRUE),]
top_crop <- df_figdata[order(df_figdata$crop_per_year, decreasing=TRUE),]
top_damage <- unique(rbind(head(top_prop, N), 
                           head(top_crop, N)))
# Plot with clustering
p <- ggplot(df_summary[df_summary$EVTYPE %in% top_damage$EVTYPE,], aes(x=prop_dmg.sum+1, y=crop_dmg.sum+1, colour=EVTYPE)) + geom_point() + stat_ellipse(geom='polygon', alpha=1/8, aes(fill=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"), limits=c(0.000001, 1E14))
p <- p + scale_x_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"), limits=c(0.000001, 1E14))
# For each EVTYPE, add a larger point to indicate the average for all years
# p <- p + geom_point(data=top_health, aes(x=fatal_per_year, y=injury_per_year)) + geom_point(aes(colour=EVTYPE))
p <- p + xlab('Annual Property Damage ($)')              # x-axis label
p <- p + ylab('Annual Crop Damage ($)')                  # y-axis label
p <- p + ggtitle("Top Weather Property Hazards")         # graph title
print(p)
```

The results indicate that Drought, Floods, and Hurricanes are the major weather-related causes of crop damage. Hurricanes, Floods, Hurricanes, and Storm Surges are major causes of property damage.

Table: Property / Crop Damage Summary

| Rank | Property Weather Events | Total / Annual Damage | Crop Weather Events | Total / Annual Damage 
| ---- | ---------- | ------------------:| ---------- | ------------------------:|
| 1 | `r top_prop$EVTYPE[1]` | $ `r top_prop$prop[1]/1E9`B / $ `r top_prop$prop_per_year[1]/1E9`B | `r top_crop$EVTYPE[1]` | $ `r top_crop$crop[1]/1E9`B / $ `r top_crop$crop_per_year[1]/1E9`B |
| 2 | `r top_prop$EVTYPE[2]` | $ `r top_prop$prop[2]/1E9`B / $ `r top_prop$prop_per_year[2]/1E9`B | `r top_crop$EVTYPE[2]` | $ `r top_crop$crop[2]/1E9`B / $ `r top_crop$crop_per_year[2]/1E9`B |
| 3 | `r top_prop$EVTYPE[3]` | $ `r top_prop$prop[3]/1E9`B / $ `r top_prop$prop_per_year[3]/1E9`B | `r top_crop$EVTYPE[3]` | $ `r top_crop$crop[3]/1E9`B / $ `r top_crop$crop_per_year[3]/1E9`B |
| 4 | `r top_prop$EVTYPE[4]` | $ `r top_prop$prop[4]/1E9`B / $ `r top_prop$prop_per_year[4]/1E9`B | `r top_crop$EVTYPE[4]` | $ `r top_crop$crop[4]/1E9`B / $ `r top_crop$crop_per_year[4]/1E9`B |
