---
title: "USA Hazardous Weather Events (1950-2011)"
author: "David Dessert"
date: "Tuesday, February 17, 2015"
output: html_document
keep_md: true
---

## Synopsis
This report identifies weather severe events most harmful to people and property. Using historical data from the [NOAA Storm Database](http://www.ncdc.noaa.gov/stormevents/) between 1950 and 2011, annualized fatalities and injuries counts of weather events were analyzed and the worst weather events identified. By our analysis, tornadoes and heat events caused the most annual fatalities and injuries while droughts, floods, and hurricanes caused the most crop and property damage.

## Downloading and Processing the Raw Source File
First, we load the necessary packages to execute
```{r}
library(ggplot2)
library(plyr)
```

This is the R Software environment we are using for this analysis.
```{r}
sessionInfo()
```

The input data file is downloaded from the Coursera web site and stored in the ./data subdirectory. 
The source file is in a comma-separated variable format (.csv extension) within 
a zipped file (.bz2 extension). Extracting and loading takes a while.
```{r}
inFile <- "./data/repdata_data_StormData.csv.bz2"
# download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
#               inFile)
data <- read.csv(bzfile(inFile))
df <- data[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
summary(df)
```

### Cleaning the Data
Extract the year of occurrence
```{r}
df$YEAR <- as.numeric(gsub("\\d+/\\d+/(\\d+).*", "\\1", data$BGN_DATE))
```

Combine EVTYPE's that are identical (ignore case difference)
```{r}
levels(df$EVTYPE) <- toupper(levels(df$EVTYPE))
```
Not all weather events were available throughout the entire database.
This table indicates the dates when weather events were available.

| Event Type(s)     | Dates Available     |
| ----------------- | ------------------- |
| Tornado           | Jan 1950 - Present  |
| Thunderstorm Wind | Jan 1955 - Present  |
| Hail              | Jan 1955 - Present  |
| All others        | Jan 1996 - Present  |

Additionally, damage estimates were not recorded in the database prior to 1993.

Here we merge some EVTYPE's into the event types enumerated by the NWS.
The people who entered the data into the database were not very consistent.
Here I coerce the various EVTYPE's used into one of the 48 EVTYPE's
spelled out specifically in the [Storm Event documentation](http://www.nws.noaa.gov/directives/).
Not all EVTYPE's remaining will have been coerced into the correct types,
but all weather events large enough to affect the analysis are corrected.

```{r EVTYPE_cleanup}
# For some reason, DEBRIS FLOW is not an entry. Add it to our levels (at the end).
levels(df$EVTYPE) <- c(levels(df$EVTYPE), "DEBRIS FLOW")

# Start coercion of EVTYPE's into the 48 "approved" types
df$EVTYPE[grep("AVAL", df$EVTYPE)] <- "AVALANCHE"
df$EVTYPE[grep("BLIZ", df$EVTYPE)] <- "BLIZZARD"
df$EVTYPE[grep("COAST.*FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("CSTL FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("BEACH FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("TIDAL FL", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COASTALSTORM", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COASTAL STORM", df$EVTYPE)] <- "COASTAL FLOOD"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("UNSEASONABLY COOL", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("HYPERTHERMIA", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("WIND CH", df$EVTYPE)] <- "COLD/WIND CHILL"
df$EVTYPE[grep("ICE.*FOG", df$EVTYPE)] <- "FREEZING FOG"
df$EVTYPE[grep("FRE.*FOG", df$EVTYPE)] <- "FREEZING FOG"
df$EVTYPE[grep("MUD.*SL", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("ROCK SLIDE", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("LANDSL", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("COASTAL EROSION", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("DAM BREAK", df$EVTYPE)] <- "DEBRIS FLOW"
df$EVTYPE[grep("DENS.*FOG", df$EVTYPE)] <- "DENSE FOG"
df$EVTYPE[grep("^FOG$", df$EVTYPE)] <- "DENSE FOG"
df$EVTYPE[grep("SMOKE", df$EVTYPE)] <- "DENSE SMOKE"
df$EVTYPE[grep("DROU", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("EXCESSIVELY DRY", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("UNSEASONABLY DRY", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("HOT/DRY PATTERN", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("WARM DRY CONDITIONS", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("RECORD DRYNESS", df$EVTYPE)] <- "DROUGHT"
df$EVTYPE[grep("DUST.*DEV", df$EVTYPE)] <- "DUST DEVIL"
df$EVTYPE[grep("DUST.*ST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("BLOWING DUST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("SAHARAN DUST", df$EVTYPE)] <- "DUST STORM"
df$EVTYPE[grep("HEAT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("UNSEASONABLY WARM & WET", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("UNSEASONABLY HOT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("CHILL", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("HYPOTHERMIA", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("LOW TEMP", df$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[grep("FLASH", df$EVTYPE)] <- "FLASH FLOOD"
df$EVTYPE[grep("^FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("FROST", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("FREEZE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("BLACK ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("PATCHY ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("GLAZE ICE", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("FUNNEL", df$EVTYPE)] <- "FUNNEL CLOUD"
df$EVTYPE[grep("ROTATING WALL CLOUD", df$EVTYPE)] <- "FUNNEL CLOUD"
df$EVTYPE[grep("HAIL", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("ICE PELLETS", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("HEAT", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("^UNSEASONABLY WARM$", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("^UNSEASONABLY WARM YEAR$", df$EVTYPE)] <- "HEAT"            # After "DROUGHT"
df$EVTYPE[grep("RAIN DAMAGE", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXTREMELY WET", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAV.* SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("RECORD.* SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOWSTORM", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOW.* RECORD", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("FIRST SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("EXCESSIVE SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("EARLY SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("^RAIN/SNOW", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SURF", df$EVTYPE)] <- "HIGH SURF"
df$EVTYPE[grep("^HIGH SEAS", df$EVTYPE)] <- "HIGH SURF"
df$EVTYPE[grep("HURRICANE", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("REMNANTS OF FLOYD", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("TYPHOON", df$EVTYPE)] <- "HURRICANE"
df$EVTYPE[grep("ICE ST", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("ICE FLOES", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("ICE JAM", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("IC[EY].* ROAD", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("LAKE FL", df$EVTYPE)] <- "LAKESHORE FLOOD"
df$EVTYPE[grep("LAKE SN", df$EVTYPE)] <- "LAKE-EFFECT SNOW"
df$EVTYPE[grep("LIGHTN", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("LIGNTNING", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("LIGHTING", df$EVTYPE)] <- "LIGHTNING"
df$EVTYPE[grep("MARINE T.* WIND", df$EVTYPE)] <- "MARINE THUNDERSTORM WIND"
df$EVTYPE[grep("ROUGH SEAS", df$EVTYPE)] <- "MARINE HIGH WIND"
df$EVTYPE[grep("MARINE MISHAP", df$EVTYPE)] <- "MARINE HIGH WIND"
df$EVTYPE[grep("RIP .*CURR", df$EVTYPE)] <- "RIP CURRENT"
df$EVTYPE[grep("SLEET", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("FREEZING RAIN", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("FREEZING DRIZZLE", df$EVTYPE)] <- "SLEET"
df$EVTYPE[grep("^HIGH WAVES", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("SURGE", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("HIGH TIDES", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("ASTRONOMICAL HIGH TIDE", df$EVTYPE)] <- "STORM SURGE/TIDE"
df$EVTYPE[grep("STR.* WIND", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND GUST", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND STORM", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("WIND ADVIS", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("GUST.*WIND", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WIND$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WINDS$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("^WIND DAMAGE$", df$EVTYPE)] <- "STRONG WIND"         # only wind, no storms
df$EVTYPE[grep("THU.*WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTM WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("DRY MICROBURST WINDS", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTM WND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TSTMW", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORMW", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORM WIN", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("THUNDERSTORM W INDS", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TUNDERSTORM WIND", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("^SEVERE THUNDERSTORM", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("GUSTNADO", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("TORN", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("WHIRLWIND", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("TROP", df$EVTYPE)] <- "TROPICAL STORM"
df$EVTYPE[grep("VOLC", df$EVTYPE)] <- "VOLCANIC ASH"
df$EVTYPE[grep("SPOU", df$EVTYPE)] <- "WATERSPOUT"
df$EVTYPE[grep("FIRE", df$EVTYPE)] <- "WILDFIRE"
df$EVTYPE[grep("WINTER S", df$EVTYPE)] <- "WINTER STORM"
df$EVTYPE[grep("WINTER W", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("SNOW SQUALLS", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("WINT.* MIX", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^ICE$", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^ICE/SNOW$", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW AND ICE", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW/BLOWING SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^LIGHT SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW/FREEZING RAIN", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^DRIFTING SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW SQUALL", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^THUNDERSNOW SHOWER", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^MIXED PRECIP", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^SNOW FREEZING RAIN", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^LAKE[ -]EFFECT SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("^MIXED PRECIPITATION", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("HEAVY RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXCESSIVE RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RAINSTORM", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RAIN \\(HEAVY\\)", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("TORR.* RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("PROLONGED RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HVY RAIN", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAVY PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HEAVY SHOWER", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("EXCESSIVE PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("RECORD PRECIP", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^THUNDERSTORM$", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^WET YEAR", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("HIGH +WIND", df$EVTYPE)] <- "HIGH WIND"         # only wind, no storms
df$EVTYPE[grep("^STORM FORCE WINDS", df$EVTYPE)] <- "HIGH WIND"         # only wind, no storms
df$EVTYPE[grep("RIVER FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("STREAM FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("URBAN.* FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("RURAL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("STREET FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HIGHWAY FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SNOW.* FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("MINOR FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("LOCAL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SMALL FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("MAJOR FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("BREAKUP FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("ICE JAM FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("SML STREAM FLD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("DAM FAILURE", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HIGH WATER", df$EVTYPE)] <- "FLOOD"
```

### Incorporate the units for Crop & Property Damages
The suffixes used in the database are quite varied. Based on a review of the
entries, here is what I'm assuming each multiplier means:  

Table: Damage Multipliers
| DMGEXP     | Multiplier         | Interpretation           |
| ---------- | ------------------ | ------------------------ |
| b, B       | 10^9^ (Billions)   | Billions                 |
| m, M       | 10^6^ (Millions)   | Millions                 |
| k, K       | 10^3^ (Thousands)  | Thousands                |
| h, H       | 10^2^ (Hundreds)   | Hundreds                 |
| 0..8       | 10^1^ (Tens)       | extra significant digit  |
| +          | 10^0^ (Ones)       | >                        |
| -          | 10^0^ (Ones)       | <                        |
| ?          | 10^0^ (Ones)       | usually 0                |
| " "        | 10^0^ (Ones)       | usually 0                |

Fix up the property and crop damage estimates using the multipliers.
```{r}
m <- rep(1, length = nrow(df))
m[toupper(df$PROPDMGEXP)=="B"] <- 1E9
m[toupper(df$PROPDMGEXP)=="M"] <- 1E6
m[toupper(df$PROPDMGEXP)=="K"] <- 1E3
m[toupper(df$PROPDMGEXP)=="H"] <- 1E2
df$PROPDMG <- df$PROPDMG * m

m <- rep(1, length = nrow(df))
m[toupper(df$CROPDMGEXP)=="B"] <- 1E9
m[toupper(df$CROPDMGEXP)=="M"] <- 1E6
m[toupper(df$CROPDMGEXP)=="K"] <- 1E3
m[toupper(df$CROPDMGEXP)=="H"] <- 1E2
df$CROPDMG <- df$CROPDMG * m
```

## Results
First, we summarize the weather events by year and event into a data frame used for the rest of the summary.
```{r}
x <- ddply(df, .(EVTYPE, YEAR), summarize, 
           count=length(EVTYPE), 
           fatal.sum=sum(FATALITIES), 
           injury.sum=sum(INJURIES), 
           prop_dmg.sum=sum(PROPDMG),
           crop_dmg.sum=sum(CROPDMG)
           )
```
For each category of fatalities, injuries, crop damage, and property damage, 
we'll be getting the top `r N` EVTYPE's for display. 
When displaying the fatalities/injuries figure, less than `r 2*N` EVTYPE's 
are displayed because of overlap in the fataility top `r N` EVTYPE's 
and injury top `r N` EVTYPE's. The same comment holds true for the
crop/property damage figure.
```{r}
N <- 4      # Number of top EVTYPE's to be plotted in each category
# Note that property damage was not reported until 1993, so the
# maximum number of years is limited to 19 (1993 - 2011)
y <- ddply(x, .(EVTYPE), summarize, 
           count=min(length(YEAR), 19),
           damage=sum(prop_dmg.sum), 
           damage_per_year=sum(prop_dmg.sum)/min(length(YEAR), 19))
top_damage <- head(y[order(y$damage_per_year, decreasing=TRUE),], 6)
```

```{r echo=FALSE, warning=FALSE, fig.width=9}
p <- ggplot(x[x$EVTYPE %in% top_damage$EVTYPE & (x$YEAR >= 1993),], aes(x=EVTYPE, y=prop_dmg.sum)) 
p <- p + geom_violin(alpha=1/4, aes(fill=EVTYPE)) 
p <- p + geom_jitter(aes(colour=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"))
p <- p + xlab('Storm Event')                        # x-axis label
p <- p + ylab('Annual Property Damage ($)')             # y-axis label
p <- p + ggtitle("Top Weather Property Hazards")      # graph title
print(p)
```

Top-N Ordered by fatalities & injuries per year
```{r echo=FALSE, warning=FALSE, fig.width=9}
y <- ddply(x, .(EVTYPE), summarize, 
           count=length(YEAR), 
           fatal=sum(fatal.sum), 
           fatal_per_year=sum(fatal.sum)/length(YEAR),
           injury=sum(injury.sum),
           injury_per_year=sum(injury.sum)/length(YEAR))
top_health <- unique(rbind(head(y[order(y$fatal_per_year, decreasing=TRUE),], N), head(y[order(y$injury_per_year, decreasing=TRUE),], N)))
# Plot with clustering
p <- ggplot(x[x$EVTYPE %in% top_health$EVTYPE,], aes(x=fatal.sum, y=injury.sum, colour=EVTYPE)) + geom_point() + stat_ellipse(geom='polygon', alpha=1/8, aes(fill=EVTYPE))
# plot the average for all years
# p <- p + geom_point(data=top_health, aes(x=fatal_per_year, y=injury_per_year)) + geom_point(aes(colour=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 10, 100, 1000, 10000), labels=c("1","10","10","1000","10K"), limits=c(0.9, 10000))
p <- p + scale_x_log10(breaks=c(1, 10, 100, 1000, 10000), labels=c("1","10","10","1000","10K"), limits=c(0.9, 10000))
p <- p + xlab('Annual Fatalities')              # x-axis label
p <- p + ylab('Annual Injuries')                # y-axis label
p <- p + ggtitle("Top Weather Health Hazards")        # graph title
print(p)
```


Top-N Ordered by property & crop damage per year
```{r echo=FALSE, warning=FALSE, fig.width=9}
y <- ddply(x, .(EVTYPE), summarize, 
           count=length(YEAR), 
           prop=sum(prop_dmg.sum), 
           prop_per_year=sum(prop_dmg.sum)/length(YEAR),
           crop=sum(crop_dmg.sum),
           crop_per_year=sum(crop_dmg.sum)/length(YEAR))
top_damage <- unique(rbind(head(y[order(y$prop_per_year, decreasing=TRUE),], N), head(y[order(y$crop_per_year, decreasing=TRUE),], N)))
# Plot with clustering
p <- ggplot(x[x$EVTYPE %in% top_damage$EVTYPE,], aes(x=prop_dmg.sum+1, y=crop_dmg.sum+1, colour=EVTYPE)) + geom_point() + stat_ellipse(geom='polygon', alpha=1/8, aes(fill=EVTYPE))
p <- p + scale_y_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"), limits=c(0.000001, 1E14))
p <- p + scale_x_log10(breaks=c(1, 1E3, 1E6, 1E9, 1E12), labels=c("1","1K","1M","1B","1T"), limits=c(0.000001, 1E14))
# plot the average for all years
# p <- p + geom_point(data=top_health, aes(x=fatal_per_year, y=injury_per_year)) + geom_point(aes(colour=EVTYPE))
p <- p + xlab('Annual Property Damage ($)')              # x-axis label
p <- p + ylab('Annual Crop Damage ($)')                # y-axis label
p <- p + ggtitle("Top Weather Property Hazards")        # graph title
print(p)
```
